---
title: "Working with Strings"
author: "KSE"
date: "1/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)

# load tidyverse package
library(tidyverse)
```

# Load data to play with

```{r}
# Load data
data(starwars)

# View data
# View(starwars)
head(starwars)

# quetions about data?
# ?starwars

```

# Detect Matches

## stringr::str_detect(string, pattern)

**1. Keep all hair colors that mention brown**

```{r}

# using just filter, no stringr
starwars %>%
  dplyr::filter(hair_color == "brown") %>%
  nrow()

# use str_detect to get more (i.e. "brown, grey")
starwars %>%
  dplyr::filter(stringr::str_detect(hair_color, "brown")) %>%
  nrow()

# Can also use grepl instead of stringr
starwars %>%
  dplyr::filter(grepl("brown", hair_color)) %>%
  nrow()

```

## stringr::str_count(string, pattern)

**2. Make a new column counting how many 'a's are in each character's name**

```{r}

starwars %>%
  dplyr::mutate(letter_a = stringr::str_count(name, "a")) %>%
  head()

```

**3. Make a new column counting how many vowels are in each character's name**

```{r}

# use | for "or" statements
starwars %>%
  dplyr::mutate(letter_a = stringr::str_count(name, "a|e|i|o|u|A|E|I|O|U")) %>%
  head()

```

# Mutate strings

## stringr::str_to_lower(string)

**4. Again, but this time change all letters to lowercase first...**

```{r}

# use | for "or" statements
starwars %>%
  dplyr::mutate(name = stringr::str_to_lower(name)) %>%
  dplyr::mutate(letter_a = stringr::str_count(name, "a|e|i|o|u")) %>%
  head()

# also can use [aeiou] which is a little easier
starwars %>%
  dplyr::mutate(name = stringr::str_to_lower(name)) %>%
  dplyr::mutate(letter_a = stringr::str_count(name, "[aeiou]")) %>%
  head()

```

*Also explore stringr::str_to_upper() and stringr::str_to_title()*

```{r}

# upper
starwars %>%
  dplyr::mutate(name = stringr::str_to_upper(name)) %>%
  head()

# title
starwars %>%
  dplyr::mutate(name = stringr::str_to_title(name)) %>%
  head()

```

## stringr::str_replace(string, pattern, replacement)

**5. Replace all mentions of "Human" with "Homo sapiens"**

```{r}

# use | for "or" statements
starwars %>%
  dplyr::mutate(species = stringr::str_replace(species, "Human", "Homo sapiens")) %>%
  head()

```

*Note: this function only replaces the FIRST instance. If you want to replace multiple, try stringr::str_replace_all(string, pattern, replacement)*

**6. Replace all "a"s with "b"s in everyone's name**

```{r}

# first use stringr::str_replace()
starwars %>%
  dplyr::mutate(name = stringr::str_replace(name, "a", "b")) %>%
  head()

# now use stringr::str_replace_all()
starwars %>%
  dplyr::mutate(name = stringr::str_replace_all(name, "a", "b")) %>%
  head()

```

**7. Replace all "brown" with "purple" hair color**

```{r}

# first use stringr::str_replace()
starwars %>%
  dplyr::mutate(hair_color = stringr::str_replace(hair_color, "brown", "purple")) %>%
  head()

```

## stringr::str_replace_na(string)

**8. Replace all NA in hair_color with "none"**

```{r}

# first use stringr::str_replace() - does this work?
starwars %>%
  dplyr::mutate(hair_color = stringr::str_replace(hair_color, NA, ".")) %>%
  head()

# now use stringr::str_replace_na()
starwars %>%
  dplyr::mutate(hair_color = stringr::str_replace_na(hair_color, ".")) %>%
  head()

```

# Subset strings

## stringr::str_extract(string, pattern)

**9. Grab the first vowel in each name**

```{r}

# use | for "or" statements
starwars %>%
  dplyr::mutate(first_vowel = stringr::str_extract(name, "[aeiou]")) %>%
  head()

```

*Can also use stringr::str_extract_all to get all instances*

## stringr::str_sub(string, start, end)

**10. Get the first three letters of eye color**

```{r}

starwars %>%
  dplyr::mutate(first_eyes = stringr::str_sub(eye_color, 1, 3)) %>%
  head()

# this can also be done with just normal string subsetting in base R
starwars %>%
  dplyr::mutate(first_eyes = substr(eye_color, 1, 3)) %>%
  head()

```

**11. Replace first three letters of eye color with "eye"**

```{r}

# This is a weird setup, but it is permanent changing this dataframe lol
stringr::str_sub(starwars$eye_color, 1, 3) <- "eye"

```

## stringr::str_subset(string, pattern)

**12. Keep only the Skywalkers**

```{r}
# This doesn't seem as helpful as a grepl or str_detect...?
# Would maybe be more useful outside of a dataframe
starwars %>%
  dplyr::filter(name %in% stringr::str_subset(name, "Skywalker")) %>%
  head()

```

## stringr::str_match(string, pattern)

**13. XXX**

```{r}
# This doesn't seem as helpful as a grepl or str_detect...?
starwars %>%
  dplyr::filter(name %in% stringr::str_subset(name, "Skywalker")) %>%
  head()

```

*Can also use stringr::str_match_all to get all instances*

# Join and Split

## stringr::str_c(string1, string2...)

**14. Give each starwars character a PhD**

```{r}

# combine ", PhD" to each name
starwars %>%
  dplyr::mutate(name = stringr::str_c(name, ", PhD")) %>%
  head()

# compare to paste
starwars %>%
  dplyr::mutate(name = paste(name, ", PhD", sep = "")) %>%
  head()

# compare to paste0
starwars %>%
  dplyr::mutate(name = paste0(name, ", PhD")) %>%
  head()

# compare to glue::glue
starwars %>%
  dplyr::mutate(name = glue::glue("{name}, PhD")) %>%
  head()

# cool glue function I just learned, glue::glue_data
glue::glue_data(starwars, "{starwars$name} has {starwars$hair_color} hair")

```

## stringr::str_split_fixed(string, pattern, n)

*My most used stringr function...*

**15. Split 'name' into 'first name' and 'last name'**

```{r}

# hint: try stringr::str_split_fixed("Luke Skywalker", " ", 2) first
stringr::str_split_fixed("Luke Skywalker", " ", 2)

# This doesn't seem as helpful as a grepl or str_detect...?
starwars %>%
  dplyr::mutate(first = stringr::str_split_fixed(name, " ", 2)[,1],
                second = stringr::str_split_fixed(name, " ", 2)[,2]) %>%
  head()

# compare this to tidyr::separate
test <- starwars %>%
  tidyr::separate(name, into = c("first", "second"), sep = " ") %>%
  head()

# sometimes separate doesn't work how I expect, usually when there are different lengths (i.e. Beru Whitesun lars)

```

## stringr::str_split(string, pattern)

*You can also just split on a pattern, useful when you have different lengths... however this is not very useful in a dataframe setting if you need to split each row into the same number of columns, so I usually don't use it*

**16. Split 'name' into 'first name' and 'last name'**

```{r}

# hint: try stringr::str_split("Luke Skywalker", " ") first
stringr::str_split("Luke Skywalker", " ")

# notice the output format is different

# This doesn't seem as helpful as a grepl or str_detect...?
starwars %>%
  dplyr::rowwise() %>%
  dplyr::mutate(first = stringr::str_split(name, " ")[[1]][1],
                second = stringr::str_split(name, " ")[[1]][2]) %>%
  head()

# If you want to make sure you are always getting the last name, not the middle name, you could try tail (but will this work for C-3PO? Nope...)
starwars %>%
  dplyr::rowwise() %>%
  dplyr::mutate(first = stringr::str_split(name, " ")[[1]][1],
                second = tail(stringr::str_split(name, " ")[[1]], n = 1)) %>%
  dplyr::mutate(second = ifelse(first == second, NA, second)) %>% # trying to get around C-3PO issue
  head()

```

# Manage length

## stringr::str_length(string)

**17. Add a new column that counts the number of letters in each name**

```{r}

# Is this accurate? Does it count spaces?
starwars %>%
  dplyr::mutate(letters = stringr::str_length(name)) %>%
  head()

```

**18. Same, but remove spaces**

```{r}

# Try to combine with str_replace?
starwars %>%
  dplyr::mutate(letters = stringr::str_length(stringr::str_replace_all(name, " ", ""))) %>%
  head()

# stringr::str_trim(string, side) could be useful, but ony trims whitespace from end or beginning... not middle
stringr::str_trim(" test ")

stringr::str_trim(" test ", "left")

```

## stringr::str_trim(string, side)

**19. Same, but remove spaces**

```{r}

# Is this accurate? Does it count spaces?
starwars %>%
  dplyr::mutate(letters = stringr::str_length(stringr::str_trim(name))) %>%
  head()

```

------------------------------------------------------------------------

# Regular Expressions

*Some people, when confronted with a problem, think "I know, I’ll use regular expressions." Now they have two problems.*

- Regular expressions are a concise language for describing patterns in strings.
- Patterns in `stringr` are interpreted as regexs
- There are some special characters (sequences of characters that have a specific meaning)
  - *i.e. `\n` = new line
- Use `writeLines()` to see how R views your string after all special characters have been parsed
- Can also use `stringr::str_view_all()` to see what the matches will be

```{r}

# check out special characters and writeLines()
writeLines("\n")

writeLines("\\n")

```

**Here are a few regular expressions you might use, check out this [cheat sheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf) or [this page](https://r4ds.had.co.nz/strings.html) for more**

#### Basics 

```{r}

# example string to practice with
s <- "C. elegans are the #1 worms!"

# match every character except a new line - pulls out the first match
stringr::str_view_all(s, ".")

# match any white space
stringr::str_view_all(s, "\\s")
stringr::str_view_all(s, "[:blank:]")

# match any non-white sapce
stringr::str_view_all(s, "\\S")

# Any numbers
stringr::str_view_all(s, "[:digit:]")

# Any letters
stringr::str_view_all(s, "[:alpha:]")

# Any letters and numbers
stringr::str_view_all(s, "[:alnum:]")

# Any punctuation
stringr::str_view_all(s, "[:punct:]")
```

#### More complicated

```{r}

# example string to practice with
s <- "C. elegans are the #1 worms!"
s2 <- "C. briggsae are the #2 worms?"
s3 <- "Tropicalis are the #3 worms..."

# match this or that
stringr::str_view_all(s, "C. elegans|C. briggsae")
stringr::str_view_all(s2, "C. elegans|C. briggsae")
stringr::str_view(c("grey", "gray"), "gr(e|a)y")

# Match start of string
stringr::str_view_all(s, "^C.")
stringr::str_view_all(s3, "^C.")

# Match end of string
stringr::str_view_all(s, "!$")
stringr::str_view_all(s2, "!$")

# To remember: If you begin with power (^) you'll end up with money ($)

# to force the entire match, you can combine ^ and &
x <- c("apple pie", "apple", "apple cake")
stringr::str_view(x, "apple")
stringr::str_view(x, "^apple$")

# Pull out a match in between two strings
stringr::str_view_all(s, "C. (.*) are")
stringr::str_match(s, "C. (.*) are")[,2]

# (.*) means to match zero or more of '.' which means any character


```

## When you use a string with stringr, it automatically wraps it in `regex()`

```{r}

# you can use comments = T to comment your regex code so you don't forget what it means!
phone <- regex("
  \\(?     # optional opening parens
  (\\d{3}) # area code
  [) -]?   # optional closing parens, space, or dash
  (\\d{3}) # another three numbers
  [ -]?    # optional space or dash
  (\\d{3}) # three more numbers
  ", comments = TRUE)

str_match("514-791-8141", phone)

```

## Miscellaneous

```{r}

# you can use apropros() to search for functions in packages currently loaded using regex
apropos("match")

# list files in directory with a specific file type
dir(pattern = "\\.Rmd$")

```


------------------------------------------------------------------------

# Biology problems

Given the following amino acid change `A371P`:

1.  Get the first amino acid

2.  Get the last amino acid

3.  Get the position of the amino acid change

Given the following info field from the *C. elegans* GFF:

`ID=gene:WBGene00269434;Name=WBGene00269434;interpolated_map_position=-26.9807;locus=sqst-5;sequence_name=T17A3.13;biotype=protein_coding;so_term_name=protein_coding_gene;curie=WB:WBGene00269434;Alias=sqst-5,T17A3.13`

4.  Get the wormbase ID

5.  Get the locus name

Given the following HTA trait names `zinc.median.TOF` and `fluoxetine.125.median.TOF`:

6.  Get the drug names (i.e. `zinc` and `fluoxetine.125`)

7.  Get the trait names (i.e. `median.TOF`)
